---
output: github_document
---
<!-- badges: start -->
[![test-coverage](https://github.com/jamestsakalos/comspat/workflows/test-coverage/badge.svg)](https://github.com/jamestsakalos/comspat/actions)
[![Codecov test coverage](https://codecov.io/gh/jamestsakalos/comspat/branch/master/graph/badge.svg)](https://app.codecov.io/gh/jamestsakalos/comspat?branch=master)
[![R-CMD-check](https://github.com/jamestsakalos/comspat/workflows/R-CMD-check/badge.svg)](https://github.com/jamestsakalos/comspat/actions)
<!-- badges: end -->

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# climenv

R functions for downloading, extracting, and plotting climatological data as a 
function of user supplied multi and single geospatial polygon and point data.

## Description

If you are a scientist seeking a convenient solution for downloading, extracting
, and plotting climatological data, consider exploring the features of this 
package. It grants you access to three widely recognised modelled data sets, 
namely WorldClim 2, CHELSA, and NASA’s SRTM. It seamlessly handles both multi 
and single geospatial polygon and point data, allowing you to extract outputs 
that can serve as covariates in various ecological studies. It also allows you 
to visualise these extractions using two common graphic options – the 
Walter-Lieth climate diagram and the Holdridge life zone classification scheme. 
The last option is a scheme of our own design which incorporates aspects of both
Walter-Leigh and Holdridge. The package's user-friendly access and extraction of
globally recognisable data sets significantly enhance its versatility and 
usability across a broad spectrum of applications.

For any questions, comments or bug reports please submit an issue here on GitHub.
Suggestions, ideas and references of new algorithms are always welcome.

## News

* June-2023: Version 1.o.0

## Main functionalities

* Downloads climate data from two main sources;
  * CHELSA
  * WorldClim 2
* Downloads elevation data from two main sources;
  * get_data()
  * elevatr
* Extracts the data;
  * as raw data for points
  * as zonal statistics for a group of points or over a spatial extent (polygon)

## Installation from the source

You can install the released version of climenv from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("climenv")
```

And the development version from [GitHub](https://github.com/jamestsakalos/climenv) with:

``` r
# install.packages("devtools")
devtools::install_github("jamestsakalos/climenv", build_vignettes = TRUE)
```
## Example

This is a basic example which shows you how to use the main climenv function:

```{r example}
library(climenv)

# Step 1. Import the Sibillini National Park Boundary
# Step 2. Run the download function
# Step 3. Run the extract function
#* See ce_download & ce_extract documentation

#' # Steps 1, 2 & 3 can be skipped by loading the extracted data
#' data(s_data)
#'
#' # Step 4. Visualise the climatic envelope using a Holdridge diagram
#'
#' plot_h(data = s_data, location_g = "High")


```

The package vignette provides detailed explanation and demonstration on
the application of climenv.

## References

Juhász-Nagy, P. (1967). On association among plant populations I. *Acta Biologica Debrecina*, 5, 43–56.

Juhász-Nagy, P. (1976). Spatial dependence of plant populations. Part 1. Equivalence analysis (an outline for a new model). *Acta Botanica Academiae Scientiarum Hungaricae*, 22, 61–78.

Juhász-Nagy, P. & Podani, J. (1983). Information theory methods for the study of spatial processes and succession. *Vegetatio*, 51, 129–140.

Juhász-Nagy, P. (1984a). Notes on diversity. Part I. Introduction. *Abstracta Botanica*,
8, 43–55.

Juhász-Nagy, P. (1984b). Spatial dependence of plant populations. Part 2. A family of new models. *Acta Botanica Hungarica*, 30, 363–402.

Juhász-Nagy, P. (1993). Notes on compositional diversity. *Hydrobiologia*, 249, 173–182.
