---
output: github_document
---
<!-- badges: start -->
[![codemeta](https://github.com/jamestsakalos/climenv/actions/workflows/codemeta.yml/badge.svg)](https://github.com/jamestsakalos/climenv/actions/workflows/codemeta.yml)
[![lint](https://github.com/jamestsakalos/climenv/actions/workflows/lint.yaml/badge.svg)](https://github.com/jamestsakalos/climenv/actions/workflows/lint.yaml)
[![pkgdown](https://github.com/jamestsakalos/climenv/actions/workflows/pkgdown.yml/badge.svg)](https://github.com/jamestsakalos/climenv/actions/workflows/pkgdown.yml)
[![R-CMD-check](https://github.com/jamestsakalos/climenv/actions/workflows/R-CMD-check.yml/badge.svg)](https://github.com/jamestsakalos/climenv/actions/workflows/R-CMD-check.yml)
[![test-coverage](https://github.com/jamestsakalos/climenv/actions/workflows/test-coverage.yaml/badge.svg)](https://github.com/jamestsakalos/climenv/actions/workflows/test-coverage.yaml)
<!-- badges: end -->

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# climenv

R functions for downloading, extracting, and plotting climatological data as a 
function of user supplied multi and single geospatial polygon and point data.

## Description

If you are a scientist seeking a convenient solution for downloading, extracting
, and plotting climatological data, consider exploring the features of this 
package. It grants you access to three widely recognised modelled data sets, 
namely WorldClim 2, CHELSA, and NASA’s SRTM. It seamlessly handles both multi 
and single geospatial polygon and point data, allowing you to extract outputs 
that can serve as covariates in various ecological studies. It also allows you 
to visualise these extractions using two common graphic options – the 
Walter-Lieth climate diagram and the Holdridge life zone classification scheme. 
The last option is a scheme of our own design which incorporates aspects of both
Walter-Leigh and Holdridge. The package's user-friendly access and extraction of
globally recognisable data sets significantly enhance its versatility and 
usability across a broad spectrum of applications.

For any questions, comments or bug reports please submit an issue here on GitHub.
Suggestions, ideas and references of new algorithms are always welcome.

## News

* June-2023: Version 1.0.0

## Main functionalities

* Downloads climate data from two main sources;
  * CHELSA, short for "Climatologies at High resolution for the Earth's Land 
  Surface Areas," encompasses a high-resolution gridded dataset containing 
  extensive climate data, including temperature and precipitation for the period
  1981–2010
  * WorldClim 2, short for "World Climate", is a comprehensive global dataset 
  offers fine-scale gridded climate data including temperature and precipitation
  for the period 1970–2000
* Downloads elevation data from two main sources;
  * SRTM, which stands for 'Shuttle Radar Topography Mission,' refers 
  specifically to the hole-filled CGIAR-SRTM dataset, providing 90-meter 
  resolution elevation data for the Earth's surface within latitudes ranging 
  from -60 to 60 degrees
  * Mapzen's synthesis digital elevation product which combines several sources 
  of digital elevation models, including SRTM, the ArcticDEM (covering all 
  areas north of 60°), EUDEM (digital elevation model over Europe)
* Extracts the data;
  * as raw data for points
  * as zonal statistics for a group of points or over a spatial extent (polygon)

## Installation from the source

You can install the released version of climenv from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("climenv")
```

And the development version from [GitHub](https://github.com/jamestsakalos/climenv) with:

``` r
# install.packages("devtools")
devtools::install_github("jamestsakalos/climenv", build_vignettes = TRUE)
```
## Example

This is a basic example which shows you how to use the main climenv function:

```{r fig1, fig.width = 16, fig.height = 8, fig.align = "center"}
library(climenv)

# Step 1. Import the Sibillini National Park Boundary
# Step 2. Run the download function
# Step 3. Run the extract function
#* See ce_download & ce_extract documentation

# Steps 1, 2 & 3 can be skipped by loading the extracted data
data("it_data")

# Step 4. Visualise the climatic envelope using a Holdridge diagram
p1 <- plot_h(data = it_data, "MED")

```


The package vignette provides detailed explanation and demonstration on
the application of climenv.

## References

Fick, S.E. and R.J. Hijmans. (2017). WorldClim 2: new 1km spatial resolution 
climate surfaces for global land areas. *International Journal of Climatology*.
37, 4302–4315. \doi{10.1002/joc.5086}

Hijmans, R.J., et al., (2023). geodata: Download Geographic Data. R package 
version 0.5-8. https://CRAN.R-project.org/package=geodata

Hollister, J. W., et al. (2021). elevatr: Access Elevation Data from Various
APIs. R package version 0.4.2. https://CRAN.R-project.org/package=elevatr

Karger, D.N., et al. (2016). CHELSA climatologies at high resolution for the 
earth’s land surface areas (Version 1.0). *World Data Center for Climate*. 
\doi{10.1594/WDCC/CHELSA_v1}

Karger, D.N., et al. (2016). CHELSA climatologies at high resolution for the
earth’s land surface areas (Version 1.1). *World Data Center for Climate*. 
\doi{10.1594/WDCC/CHELSA_v1_1}

Karger, D.N., et al. (2018). Data from: Climatologies at high resolution for the
earth’s land surface areas. *EnviDat*. \doi{10.16904/envidat.228.v2.1}
