---
title: "Getting started with 'climenv'"
author: "James L. Tsakalos (james.tsakalos@gmail.com), Martin R. Smith, Federico Luebert, Ladislav Mucina"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting started with 'climenv'}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{=html}
<style>
body {
text-align: justify}
</style>
<style>
p.caption {
  font-size: 0.9em;
  font-style: normal;
  color: grey;
  margin-right: 10%;
  margin-left: 10%;  
  text-align: justify;
}
</style>
```

------------------------------------------------------------------------

```{r knitr-setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## An example application of climenv

This vignette shows on example application of `climenv` (short for 
***Clim***atic ***Env***elope) to illustrate the concepts and functionality
incorporated in the package.
Specifically, the example facilitates easy downloading, extraction
and visualization of three of the most globally recognizable modeled data sets
(CHELSA, WorldClim 2 and NASA’s SRTM elevation) using example multi and single
geospatial polygon and point data.

Specifically, supplied with geospatial polygon and point data the package can
easily download, extract and visualize climatic data. The visualization options
include a custom plot coupled with helpful tabulated climatic properties,
Holdridge's (1967) life zone classification and Walter-Lieth's (1960)
climate diagrams. To our knowledge this is the first time that the Holdridge 
diagram has been presented in R.

### Downloading climate and elevation data

`ce_download` downloads WorldClim 2 (Fick & Hijmans, 2017), CHELSA (Karger et 
al., 2017) and NASA Earth Explorer's SRTM (Farr et al., 2007). CHESLA and 
WorldClim are available at a spatial resolution of 30 arc-seconds (c. 1 km<sup>2</sup>). The 
data are freely available as a series of raster tiles (one for each month) with 
their spatial extent spanning the globe. Specifically, the function downloads 
the mean, minimum and maximum temperature and mean precipitation using the 
climatic predictions for periods 1979–2013 (CHELSA) and 1970–2000 (WorldClim 2). 

`ce_download` also downloads NASA's SRTM data at the same 30 arc-second
(c. 1 km<sup>2</sup>) resolution. However, unlike CHELSA and WorldClim, the SRTM data can be
downloaded as a series of individual tiles. While the spatial extent of CHELSA
and WorldClim 2 spans the globe, `ce_download` derives and mosaics the SRTM
tiles matching the extent provided by the user. Downloading a subset of tiles
is advantageous because it will download quicker and occupy less storage.

Because the vignette is required to run quickly and downloading climatic data
(4.67 or 13.7&nbsp;GBs for CHELSA and WorldClim 2, resp.) takes considerable time
(ca. 6 or 19 minutes at 100&nbsp;Mbit/s). For simplicity, we have provided code to
create artificial climate data of known properties. For a full demonstration on
how to download data please see the package help documentation or Tsakalos
et al. (2021). 

```{r load-libraries, echo=FALSE, message=FALSE}
library("climenv")
library("fs")
library("raster")
library("dplyr")
library("sf")
library("terra")
```

```{r training-data}
# Let's make some training data

# Create temporary file to supply to the ce_extract
temp_path <- tempfile()

# Create the  sub-folders
dir.create(file.path(temp_path, "elev"), recursive = TRUE)
dir.create(file.path(temp_path, "prec"), recursive = TRUE)
dir.create(file.path(temp_path, "tmax"), recursive = TRUE)
dir.create(file.path(temp_path, "tavg"), recursive = TRUE)
dir.create(file.path(temp_path, "tmin"), recursive = TRUE)
on.exit(unlink(file.path(temp_path)), add = TRUE)

# Create a empty raster serving as a base
r <- terra::rast(ncol = 10, nrow = 10)

# Modify the base Raster values and save them in correct configuration

#* Elevation 100m ####
terra::values(r) <- 1:100
terra::writeRaster(r, paste0(temp_path, "/elev/srtm.tif"))

# Prec
x <- c(5, 10, 15, 20, 25, 34.40666, 25, 20, 15, 10, 5, 0) * 8
temp2 <- paste0("prec_", sprintf("%02d", 1:12), ".tif")
for (i in seq_along(temp2)) {
  terra::values(r) <- x[i]
  terra::writeRaster(r, paste0(temp_path, "/prec/", temp2[i]))
}

# tmax
x <- c(43, 38, 33, 29, 25, 19.8, 17.01, 21, 25, 30, 37, 44)
temp2 <- paste0("tmax_", sprintf("%02d", 1:12), ".tif")
for (i in seq_along(temp2)) {
  values(r) <- x[i]
  writeRaster(r, paste0(temp_path, "/tmax/", temp2[i]))
}

# tmin
x <- c(43, 38, 33, 29, 25, 19.8, 17.01, 21, 25, 30, 37, 44) / 2
temp2 <- paste0("tmin_", sprintf("%02d", 1:12), ".tif")
for (i in seq_along(temp2)) {
  values(r) <- x[i]
  writeRaster(r, paste0(temp_path, "/tmin/", temp2[i]))
}

# tmean
x <- c(43, 38, 33, 29, 25, 19.8, 17.01, 21, 25, 30, 37, 44) -
  c(c(43, 38, 33, 29, 25, 19.8, 17.01, 21, 25, 30, 37, 44) / 2) / 2
temp2 <- paste0("tavg_", sprintf("%02d", 1:12), ".tif")
for (i in seq_along(temp2)) {
  values(r) <- x[i]
  writeRaster(r, paste0(temp_path, "/tavg/", temp2[i]))
}

# Create a polygon file from the raster
terra::values(r) <- 1:100
pol_py <- sf::st_as_sf(terra::as.polygons(r))
pol_py$grp <- c(rep("low", 25), rep("high", 75))

# Create a point file from the raster
pol_pt <- sf::st_as_sf(terra::as.points(r))
pol_pt$grp <- c(rep("low", 25), rep("high", 75))

```

### Extracting the statistics of the distribution of values for each climatic variable within the area

Now that the climate and elevation data have been created, the next step is to
use `ce_extract` to extract the climatic data using the pol_py and 
pol_pt geospatial data sets. This function's general workflow is to read in the
newly downloaded data as a series of raster stacks (processing time should be
shorter when using a raster stack, see Hijmans, 2020). Once the stacks are read
in, the function continues to crop and mask using a unique feature stored in the
multi-point or multi-polygon geospatial data set. For example, the climatic
and environmental data can be extracted for all Sibillini National Park or high
and low sections. This feature is controlled by the 'location.g' argument,
specifying a column in the geospatial data set.

`ce_extract` adds a buffer of 2 km when small polygon sizes return no
results. The polygon is transformed from a geographic coordinate system into
a projected coordinate system using the 'spTransform' function. This
transformation allows for a metric buffer to be added to the polygon.

```{r extract-data}
# Extract the climate data

data_py <- ce_extract(
  file.path(temp_path),
  location = pol_py,
  location_g = NULL
)

data_pt <- ce_extract(
  file.path(temp_path),
  location = pol_pt,
  location_g = NULL
)
```

The returned data object is a list containing five data sets. Specifically, one 
data set each is returned for precipitation and  mean-, maximum- and 
minimum-temperature. One data set is produced for altitude and continentality. 
These data sets are amenable to further use by the user, such as covariates in 
any number of modelling exercises. 

### Visualizing the climatic and elevation data

`climenv` offers three functions for visualising the climatic data extracted 
from points or polygons in various formats.

The third climate diagram is one of 
our own design (see Macintyre & Mucina, 2021).

`plot_h` produces Holdridge's life zone classification plot, also known as the 
Holdridge Life Zone System or Holdridge Bioclimatic Classification System, is 
based on three main factors that influence the distribution of vegetation and 
ecosystems. By combining temperature, precipitation, and potential 
evapotranspiration Holdridge's classification plot divides the Earth’s surface
into distinct life zones or biomes (sensu Holdridge). It allows for the 
identification and characterization of different biomes, such as tropical 
rainforests, deserts, grasslands, and tundra, based on their distinct climatic 
conditions and provides a unified framework for studying vegetation patterns, 
ecological dynamics, and potential shifts in response to climate change. To 
simplify the visualization of life zone data, we have implemented the automatic 
creation of Holdridge plots by the addition of the `plot_h` function which 
provides a convenient wrapper from within `climenv` for the function 
`PlotHoldridge` within the `Ternary` R package (Smith 2017). This paper is the 
formal description of this plotting tool which has been developed to complement 
`climenv`

`plot_wl` produces the Walter-Lieth (1960) climatic diagram. This diagram
consists of two primary components: temperature and precipitation, which when 
combined in a single diagram allows for a comprehensive visualisation of 
climate patterns. Specifically, it provides insights into seasonal variations, 
the duration and intensity of wet and dry periods, and the overall climate 
regime of a particular location (or the average for an area encompassed by a 
spatial polygon). Here our package is a wrapper for the existing `diagwl` 
function of the `climatol` R package (Guijarro, 2019).


```{r holdridge-diagram, fig.height = 5, fig.width = 5, fig.align = "center", fig.cap="Fig 1. Position of the training data within Holdridge's (1967) life zone classification. The surface shading in the background is new addition to the original life zone classification and helps interpretation by converting a point in evapotranspiration-precipitation space to an appropriate cross-blended hypsometric colour – in this intuitive instance colours tending towards the red spectrum feature higher temperatures blended with lower precipitation compared while colours tending towards the blue colour spectrum have lower temperatures and higher precipitation."}
# Make Holdridge's (1967) life zone classification diagram
plot_h(data = data_py, "0")
```
```{r walter-leigh-diagram, fig.height = 5, fig.width = 5, fig.align = "center", fig.cap="Fig 2. Walter-Lieth’s climatic diagram (1960) of the training data. When precipitation is > 100 mm, the scale increases from 2mm C-1 to 20 mm  C-1 (as indicated by the black horizontal line) to avoid too high diagrams in very wet locations. This change is indicated by a black horizontal line, and the graph over is filled in solid blue. When the precipitation graph lies under the temperature graph (P < 2T) we have an arid period (filled in dotted red vertical lines). Otherwise the period is considered humid (filled in light blue). Daily maximum average temperature of the hottest month and daily minimum average temperature of the coldest month are labeled in black on the left margin of the diagram."}
# Make Walter-Leigh's (1960) climate diagram
plot_wl(data = data_py, "0")
```
```{r custom-climate-diagram, fig.height = 4, fig.width = 7, fig.align = "center", fig.cap="Fig 3. Custom diagram showing the climatic envelope of the training data. The abbreviations used are as follow: biotemperture (BioT), isothermality (ISO), mean annual temperature (MAT), temperature seasonality (TS), number of dry months with < 50 mm rainfall during the month (Dry mo), mean annual precipitation (MAP), potential evapotranspiration (PET), precipitation seasonality (PS), seasonal rainfall percentage in Summer (S), Autumn (A), Winter (W), Vernal (V), elevation (Elv) and latitude (Lat)."}
# Make the custom climate diagram
oldpar <- par(mar = c(1.5, 2.8, 2, 17))
on.exit(par(oldpar))
plot_c(data = data_py, "0")
```
