---
title: "Exploring the Italian biome data sets"
author: "James L. Tsakalos (james.tsakalos@gmail.com), Martin R. Smith, Frederico Luebert, Ladislav Mucina"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Exploring the Italian biome data sets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{=html}
<style>
body {
text-align: justify}
</style>
```

------------------------------------------------------------------------

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## General description of the Italian Biome data sets

`climenv` includes three data sets that feature Biome information for Italy. The
data sets include `it_py`, `it_pt` and `it_data`. The prefix it stands for 
"Italy". Within `it_py` the Biomes are represented by polygons (py), within 
`it_pt` the Biomes have been sampled by a gridded (~5km) point (pt) data set. 
Lastly, `it_data` is the output of the `ce_extract()` function generated using
WorldClim 2 climate and Mapzen digital elevation.

```{r, message = FALSE}
library(climenv)
library(terra)

#library(fs)
#library(raster)
#library(dplyr)
#library(sf)

# Lets begin by visualising the data

data("it_py")
plot(it_py, key.pos = NULL, border= NA, main = "Global Biomes polygons")

data("it_pt")
plot(it_pt, pch = 19, cex = 0.5, key.pos = NULL, main = "Global Biomes points")

# Due to downloading time we have provided the extracted data in the package
## ce_download("C/example_output..", c_source = "WorldClim", location = it_py)
## ?ce_extract("C/example_output", "C/example_output/elev", it_py, "GB")

# Lets have a look at the extracted data

data("it_data")
str(it_data)

```

