---
title: "Exploring the Italian biome data sets"
author: "James L. Tsakalos (james.tsakalos@gmail.com), Martin R. Smith, Federico Luebert, Ladislav Mucina"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Exploring the Italian biome data sets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{=html}
<style>
body {
text-align: justify}
</style>
<style>
p.caption {
  font-size: 0.9em;
  font-style: normal;
  color: grey;
  margin-right: 10%;
  margin-left: 10%;  
  text-align: justify;
}
</style>
```

------------------------------------------------------------------------

```{r knitr-setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Spatial distribution of the Italian Biome data sets

`climenv` includes three data sets that feature Biome information for Italy. The
data sets include `it_py`, `it_pt` and `it_data`. The prefix it stands for 
"Italy". Within `it_py` the Biomes are represented by polygons (py), within 
`it_pt` the Biomes have been sampled by a gridded (c. 5km) point (pt) data set. 
Lastly, `it_data` is the output of the `ce_extract()` function generated using
WorldClim 2 climate and Mapzen digital elevation.

```{r initialize, message = FALSE}
library("climenv")
library("terra")

# Lets begin by loading the data

data("it_py")
data("it_pt")
```
```{r italian-polygons-map, fig.height = 5, fig.width = 5, fig.align = "center", fig.cap = "**Fig 1**. Mediterranean and Nemoral Biomes of Italy (Mucina et al., 2023). The Mediterranean Biome is shaded green, while the Nemoral Biome is shaded in orange."}
# Make a map of the Italian polygons data set
plot(it_py, key.pos = NULL, border = NA, main = "")
```

From **Fig 1**, it is evident that the Italian polygons data set effectively 
represents the spatial distribution of the Mediterranean and Nemoral Biomes 
derived from the aggregation of vegetation typologies across the entire 
geography of Italy. This data set offers valuable insights into the extent and 
boundaries of different Biomes within the region. Nevertheless, it is essential 
for users to recognise that should they have smaller or larger scaled maps, 
which capture the spatial boundaries of different vegetation groups such as 
communities, habitats, disturbance classes, etc., they would serve as an ideal 
candidate for analyses with our package.

```{r italian-points-map, fig.height = 5, fig.width = 5, fig.align = "center", fig.cap = "**Fig 2**. Mediterranean and Nemoral Biomes of Italy (Mucina et al., 2023). The Mediterranean Biome is coded as green points, while the Nemoral Biome is coded as orange points."}
# Make a map of the Italian points data set
plot(it_pt, pch = 19, cex = 0.5, key.pos = NULL, main = "")
```

By design **Fig 2**, spans almost the same spatial extent as in **Fig 1**. Each 
of the points (n = 9963) within the Italian point data set is spaced at c. 5m in 
a grid configuration. Here users containing georeferenced features such as 
quadrat locations, mapping notes of vegetation typologies, disturbance regimes, 
etc., they would serve as an ideal candidate for analyses with our package. By 
utilising these polygon or point mapping structures as a foundation, our package
enables users to leverage the inherent structure and spatial representation they
require to conduct their own targeted investigations effectively. This 
flexibility accommodates a wide range of research needs and empowers users to 
explore the intricate relationships of vegetation groups, habitats, disturbance 
classes, structural classes, etc., on their terms.

### Downloading climate and elevation data

The function `ce_download` enables users to download both climate and elevation 
data. Users have the option to select climate data from either WorldClim 2 
(Fick & Hijmans, 2017) or CHELSA (Karger et al., 2017; Karger et al., 2021). 
Regarding elevation data, users can choose either the NASA Earth Explorer's SRTM
(Farr et al., 2007) or Mapzen terrain tiles (Hollister & Shah, 2018).

In this specific example, we have utilised the default settings  of 
`ce_download` to download WorldClim 2 climate data and Mapzen elevation data for 
the Italian polygon data set. However, please note that the following code chunk 
is intentionally set not to evaluate because its execution requires time. 
Fortunately, to facilitate the presented examples, we have already stored the 
downloaded and extracted (using `ce_extract`) data within the package 
(see below for more details).

```{r extract-data, message = FALSE, eval = FALSE}
# Do not execute this code if you are in a hurry as it will take time depending
# on the strength and speed of your internet connection.

# This part downloads the data and takes time
ce_download(
  "C/example_output..",
  location = it_py
)

# Once the data is downloaded, this function is quick.
it_data <- ce_extract(
  "C/example_output..",
  location = it_py,
  location_g = "GB"
)
```
```{r load-data, message = FALSE}
# Fortunately you can avoid the above steps by simply loading the data
data("it_data")
```

### Climatic envelope of the Italian Biome data sets

climenv provides three plotting functions, one for the traditional Holdridge 
life zone classification scheme (`plot_h`), one for the Walter-Lieth climate 
diagrams (`plot_wl`) and a new custom plot (`plot_c`) integrating features from 
both systems as well as the commonly used bioclim variables (Hijmans et al., 
2005).

One of our graphical outputs is the Holdridge (1967) life zone classification
plot. Holdridge's life zone classification plot, also known as the Holdridge 
Life Zone System or Holdridge Bioclimatic Classification System, is based on 
three main factors that influence the distribution of vegetation and ecosystems.
By combining temperature, precipitation, and potential evapotranspiration 
Holdridge's classification plot divides the Earth’s surface into distinct life 
zones or biomes (*sensu* Holdridge).

```{r holdridge-diagram, fig.show = 'hold', fig.height = 5, fig.width = 5, fig.align = "center", fig.cap = "**Fig 3**. Position of the Mediterranean (MED) and Nemoral (NEM) Biomes within Italy derived using WorldClim climate within Holdridge's (1967) life zone classification. The surface shading in the background is new addition to the original life zone classification and helps interpretation by converting a point in evapotranspiration-precipitation space to an appropriate cross-blended hypsometric colour – in this intuitive instance colours tending towards the red spectrum feature higher temperatures blended with lower precipitation compared while colours tending towards the blue colour spectrum have lower temperatures and higher precipitation."}
# Make Holdridge's (1967) life zone classification diagram
plot_h(data = it_data, "MED")
plot_h(data = it_data, "NEM")
```

Another common graphical output is the Walter-Lieth (1960) climatic diagram. 
Here our package is a wrapper for the existing diagwl function of the climatol 
R package (Guijarro, 2019). This diagram consists of two primary components: 
temperature and precipitation, which when combined in a single diagram is 
supposed to allow for a comprehensive visualisation of climate patterns.

```{r walter-leigh-diagram, fig.show = 'hold', fig.height = 5, fig.width = 5, fig.align = "center", fig.cap = "**Fig 4**. Walter-Lieth’s climatic diagram (1960) of the Mediterranean (MED) and Nemoral (NEM) Biome within Italy derived using WorldClim. When precipitation is > 100 mm, the scale increases from 2mm C-1 to 20 mm  C-1 (as indicated by the black horizontal line) to avoid too high diagrams in very wet locations. This change is indicated by a black horizontal line, and the graph over is filled in solid blue. When the precipitation graph lies under the temperature graph (P < 2T) we have an arid period (filled in dotted red vertical lines). Otherwise the period is considered humid (filled in light blue). Daily maximum average temperature of the hottest month and daily minimum average temperature of the coldest month are labeled in black on the left margin of the diagram."}
# Make Walter-Leigh's (1960) climate diagram
plot_wl(data = it_data, "MED")
plot_wl(data = it_data, "NEM")
```

Lastly, we present our custom diagrams which incorporate elements from 
Holdridge's (1967) life zone classification, Walter-Lieth's climatic diagram 
(1960), and the widely utilised bioclimatic variables (Hijmans et al., 2005) 
commonly employed in ecological and environmental research. These variables are 
crucial in describing environmental factors that shape the distribution and 
behaviour of organisms, including plants, as evidenced by the high citation
count of works by Holdridge, Walter-Lieth, and Hijmans. 

```{r custom-climate-diagram, fig.show = 'hold', fig.height = 4, fig.width = 7, fig.align = "center", fig.cap = "**Fig 5**. Custom diagram showing the climatic envelope of the Italian Mediterranean (MED) and Nemoral (NEM) Biomes. The abbreviations used are as follow: biotemperture (BioT), isothermality (ISO), mean annual temperature (MAT), temperature seasonality (TS), number of dry months with < 50 mm rainfall during the month (Dry mo), mean annual precipitation (MAP), potential evapotranspiration (PET), precipitation seasonality (PS), seasonal rainfall percentage in Summer (S), Autumn (A), Winter (W), Vernal (V), elevation (Elv) and latitude (Lat)."}
# Make the custom climate diagram
par(mar = c(1.5, 2.8, 2, 17))
plot_c(data = it_data, "MED")
plot_c(data = it_data, "NEM")
```

### Notes on selecting appropriate climatic variables

To ensure the selection of climatic variables for discriminatory analyses, 
numerous data-driven approaches can be employed. For instance, with lots of 
data, one could use CART, random forests, boosted regression trees, or any 
number of machine learning algorithms to determine which variables are the most 
appropriate (robust and ecologically meaningful) for delineating between classes
, for the purposes of discriminatory analyses. In this extended vignette, we 
will delve into an exploration of which climatic variables are most appropriate 
for delineating our highly specific example data set of Italian Biomes (Italian 
point data set). By employing these rigorous methods, one can make a more 
informed decisions about the choice of climatic variables that play a major role
in characterising and distinguishing the climatic envelopes of the various 
biomes in their own study areas.

```{r extract-extended-data, message = FALSE, eval = FALSE}
# Lets start by extracting the climate data for all points in the
# italy_pt data.

it_data_extended <- ce_extract(
  path = "C:/Users/jamie/OneDrive/Desktop/temp",
  location = it_pt
)
```
```{r load-extended-data, message = FALSE}
# If you haven't downloaded the climate and elevation data by following the code
# above you could simply load the data()

data("it_data_extended")

# But this data needs to be reformatted a bit

# Lets start by calculating the bioclim variables

bioclim <- data.frame(dismo::biovars(
  prec = as.matrix(it_data_extended$prec[, 1:12]),
  tmin = as.matrix(it_data_extended$tmin[, 1:12]),
  tmax = as.matrix(it_data_extended$tmax[, 1:12])
))

# Lets make some more intuitive names for these variables
colnames(bioclim) <- c(
  "Mean_annual_temperature",
  "Mean_diurnal_range",
  "Isothermality",
  "Temperature_seasonality",
  "Max_temperature_of_warmest_month",
  "Min_temperature_of_coldest_month",
  "Temperature_annual_range",
  "Mean_temperature_of_the_wettest_quarter",
  "Mean_temperature_of_driest_quarter",
  "Mean_temperature_of_warmest_quarter",
  "Mean_temperature_of_coldest_quarter",
  "Total_annual_precipitation",
  "Precipitation_of_wettest_month",
  "Precipitation_of_driest_month",
  "Precipitation_seasonality",
  "Precipitation_of_wettest_quarter",
  "Precipitation_of_driest_quarter",
  "Precipitation_of_warmest_quarter",
  "Precipitation_of_coldest_quarter"
)

# Now get the Holdridge variables

hold <- bioclimate(
  as.matrix(it_data_extended$tavg[, 1:12]),
  as.matrix(it_data_extended$prec[, 1:12])
)

# Make some more intuitive names for these variables
colnames(hold)[1:3] <- c(
  "Mean_Annual_Biotemperature",
  "Total_Annual_Precipitation",
  "Potential_Evapotranspiration_Ratio"
)

# Join them together along with elevation and altitude

dat <- cbind(
  bioclim, hold[, c("Mean_Annual_Biotemperature",
                    "Potential_Evapotranspiration_Ratio")],
  elev = it_data_extended$elev,
  lat = it_data_extended$lat
)

# Lets just fix the last two column names
colnames(dat)[c(22, 23)] <- c("Elevation", "Latitude")

# Now lets add the Biome type as a column

dat$class <- as.factor(it_pt$GB)

```

Random Forest is an ensemble learning algorithm that combines multiple decision 
trees to make predictions. Decision trees are structures that make predictions 
by splitting data based on if-else conditions. They create a tree-like model of 
decisions. Random Forest adds randomness by using a technique called 
bootstrapping, which involves randomly sampling the data with replacement. It 
also introduces variability by randomly selecting a subset of predictor 
variables for each split in a decision tree. In Random Forest, a collection of 
decision trees is built. Each tree is trained on a different subset of the data 
and uses a random subset of predictor variables. Random Forest models are known 
for their robustness, accuracy, and resistance to overfitting. They handle 
high-dimensional data well, are less affected by outliers, and can capture 
complex relationships between variables. Additionally, Random Forest provides an
estimate of variable importance, which helps with feature selection and 
interpretation. For more information please refer to Breiman (2001).

```{r random-forests, message = FALSE}
# Now we are about ready to start the random forests analyses

# Load the required package
library("randomForest")

# Now we need to split the dataset into training and testing sets
set.seed(123)  # For reproducibility

train_indices <- sample(
  seq_len(dim(dat)[1]), dim(dat)[1] * 0.8
)  # 80% for training
train_data <- dat[train_indices, ]
test_data <- dat[-train_indices, ]

# Train the Random Forest model
model <- randomForest(formula = class ~ ., data = train_data)

# Make predictions on the test set
predictions <- predict(model, newdata = test_data)

# Evaluate the model's performance
confusion_matrix <- table(predictions, test_data$class)
accuracy <- sum(diag(confusion_matrix)) / sum(confusion_matrix)
print(confusion_matrix)
print(paste("Accuracy:", accuracy))
```

Based on the provided results, the model demonstrated a high accuracy of 
approximately 95% in predicting the Mediterranean and Nemoral Biomes in Italy. 
The individual error rates were observed to be around 8% for the Mediterranean
Biome and 4% for the Nemoral Biome. While cross-validation and other methods 
could potentially enhance the model's performance, exploring them falls beyond 
the scope of this brief vignette. However, it might be valuable to assess the 
importance of variables in this model and determine whether they correspond to 
the variables present in our climate diagrams.

Considering this objective, the Gini coefficient is commonly employed to assess 
the performance of the predictive variables, specifically climate and elevation 
variables. The Gini coefficient, ranging from 0 to 1, serves as a measure of 
the relative importance of these predictor variables within the model. Higher 
Gini coefficients signify greater significance of the corresponding variables 
in predicting the desired outcome. By visualising the variable importance 
rankings, the plot sheds light on the crucial predictors that contribute to the 
model's performance. Variables with higher Gini coefficients hold more influence
within the Random Forest model, emphasising their importance in the prediction 
process.

```{r importance-measures, message = FALSE, fig.height = 5, fig.width = 7, fig.align = "center", fig.cap = "**Fig 6**. Gini Coefficient Plot illustrating the variable importance measures derived from a Random Forest model."}
# Get variable importance measures
importance <- importance(model)

# Print the importance measures
print(importance)

# Plot variable importance
par(mar = c(4, 3, 0, 1) + 0.1)
varImpPlot(model, main = "")
```

### References

Breiman, L. (2001). Random forests. *Machine learning*, 45, 5&ndash;32.

Farr, T. G., Rosen, P. A., Caro, E., Crippen, R., Duren, R., Hensley, S. et al. 
(2007). The Shuttle Radar Topography Mission. *Reviews of Geophysics*, 45. \doi{10.1029/2005RG000183}

Fick, S. E., & Hijmans, R. J. (2017). WorldClim 2: new 1‐km spatial resolution
climate surfaces for global land areas. *International Journal of Climatology*,
37, 4302&ndash;4315.

Guijarro, J. A. (2019). climatol: Climate Tools (Series Homogenization and 
Derived Products). Retrieved from https://cran.r-project.org/package=climatol

Hijmans, R. J., Cameron, S. E., Parra, J. L., Jones, P. G., & Jarvis, A. (2005).
Very high resolution interpolated climate surfaces for global land areas. 
*International Journal of Climatology*, 25, 1965–1978. \doi{10.1002/joc.1276}

Holdridge (1967), _[Life zone ecology]_. Tropical Science Center, San 
Jos&eacute;: Tropical Science Center.
[Life zone ecology]: https://reddcr.go.cr/sites/default/files/centro-de-documentacion/holdridge_1966_-_life_zone_ecology.pdf

Hollister, J., & Shah, T. (2018). elevatr: Access elevation data from various 
APIs (Version 0.2.0). Retrieved from https://CRAN.R-project.org/package=elevatr.
\doi{10.5281/zenodo.400259}

Karger,  D.N., Conrad, O., B&odblac;hner, J., Kawohl, T., Kreft, H., Soria-Auza,
R.W. et al. (2017) Climatologies at high resolution for the earth’s land surface 
areas. *Scientific Data*, 4, 170122. \doi{10.1038/sdata.2017.122}

Karger, D.N., Conrad, O., B&odblac;hner, J., Kawohl, T., Kreft, H., Soria-Auza, 
R.W. et al. (2021) Climatologies at high resolution for the earth’s land surface 
areas. *EnviDat*. \doi{10.16904/envidat.228.v2.1}

Mucina, L., Div&iacute;&scaron;ek, J., & Tsakalos, J.L. (2023) Europe, 
Ecosystems of. In: Encyclopedia of biodiversity, vol X (in print). 
\doi{10.1016/B978-0-12-822562-2.00059-1}

von Walter, H.B., & Lieth, H. (1960). Klimadiagramm-Weltatlas. VEB Gustav 
Fischer Verlag, Jena.
